---
title: "Application of Central Limit Theorem to a exponential distribution"
author: "Eduardo Bortoluzzi Junior"
date: "October 17, 2015"
output: pdf_document
---

Overview
--------

This report simulates an exponential distribution and a distribution of averages
of exponentials to show their differences and their similarities. It is expected
that the mean of both distributions are almost the same, but the observations
are more dispersed on the exponential distribution and more concentrated at the 
distribution of averages of exponentials. Also, the distribution of averages of
exponentials must be a standard normal. 

Simulations
-----------
<!-- 
Include English explanations of the simulations you ran, with the accompanying R code. Your explanations should make clear what the R code accomplishes. 
-->

```{r echo=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
```

Start setting some constants for the simulation.

```{r}
lambda <- 0.2

dist.mean <- 1 / lambda
dist.sd <- 1 / lambda

observations <- 40

simul <- 1000
```

$\lambda$, which is the ratio in an exponential distribution, is set to 
`r lambda` for all simulations. The distribution mean, $\mu$, is set to 
$\frac{1}{\lambda}$ in the variable `dist.mean`, which value is `r dist.mean`. 
The standard deviation, $\sigma$, is also set to $\frac{1}{\lambda}$, in the 
variable `dist.sd`, which value is `r dist.sd`.

The investigation will be on the distribution of averages of `r observations` 
exponentials, which is set in the variable `observations`. `r simul` simulations
will be run. This value is set in the variable `simul`.

To make this simulation reproducible, the seed is set:

```{r}
set.seed(20151017)
```

Two distributions are generated: an exponential distribution with 
`r simul` observations and a distribution of `r simul` averages of 
`r observations` exponentials. Each of the observations receives a `type` that
defines from which distribution it is.

```{r}
distributions <- data.frame(
    x = c (
        # exponential distribution with `simul` observations
        rexp(simul, lambda),
        
        # distribution of `simul` averages of `observations` exponentials
        replicate(simul, {
          mean(rexp(observations, lambda))
        })
    ),
    type = factor( rep(
        c('exp', 'avg'), 
        rep(simul, 2)
    ))
)

```

The sample and the standard deviation of each distribution are calculated and
saved to the variable `s`:

```{r}
s <- summarise(
    group_by(distributions, type), 
    mean=mean(x), 
    var=sd(x)^2
)
```

Recall that the $Var(\bar{X})=\frac{\sigma^2}{n}$, and the standard error 
($SE_{mean}$) is $\frac{\sigma}{\sqrt{n}}$, so $Var(\bar{X})={SE_{mean}}^2$.

A standard normal distribution, with the same paremeters from the distribution
of the averages, is generated to make a comparision:

```{r}
std.normal <- data.frame(x=rnorm(
                            n = simul, 
                            mean = s[s$type == "avg",]$mean, 
                            sd = sqrt(s[s$type == "avg",]$var)
                         ))
```


Sample Mean versus Theoretical Mean
-----------------------------------
<!--
Include figures with titles. In the figures, highlight the means you are comparing. Include text that explains the figures and what is shown on them, and provides appropriate numbers.
-->

```{r mean, echo=FALSE, fig.height=3}
g <- ggplot(distributions, aes(x=x, fill=type))

g <- g + geom_histogram(alpha=.2, 
                        binwidth=.3, 
                        aes(y=..density..), 
                        colour="black", 
                        position="identity")

g <- g + geom_vline(data=s, 
                    aes(xintercept=mean, colour=type), 
                    linetype=c("dashed","dotdash"), 
                    size=1)

g <- g + theme(legend.justification=c(1,1), legend.position=c(1,1))

g <- g + ggtitle("Exponential Distribution and Distribution of Averages")

g <- g + scale_fill_discrete(
            breaks=c("exp", "avg"), 
            labels=c("of exponentials", "of averages")
        )
g
```

The mean of the exponential distribution and the distribution of averages are
almost the same, `r s[s$type=='exp',]$mean` and 
`r s[s$type=='avg',]$mean`, respectively. The expected mean
is `r dist.mean`, so the distributions' means are what we were expecting.

Important to notice that the observations in the exponential distribution are
more dispersed and the averages of the observations in the distribution of
averages are more concentrated, but both have the same mean. This observation
is expected, in accordance with the Central Limit Theorem.

Sample Variance versus Theoretical Variance
-------------------------------------------
<!--
Include figures (output from R) with titles. Highlight the variances you are comparing. Include text that explains your understanding of the differences of the variances.
-->

```{r variance, echo = FALSE, fig.height=3}
g <- ggplot(distributions, aes(x=x, fill=type))

g <- g + geom_histogram(alpha=.2, 
                        binwidth=.3, 
                        aes(y=..density..), 
                        colour="black", 
                        position="identity")

g <- g + geom_vline(data=s, 
                    aes(xintercept=mean, colour=type), 
                    linetype=c("dashed","dotdash"), 
                    size=1)

g <- g + geom_vline(data=s, 
                    aes(xintercept=-sqrt(var)+mean, colour=type), 
                    alpha=.8,
                    size=1)

g <- g + geom_vline(data=s, 
                    aes(xintercept=+sqrt(var)+mean, colour=type), 
                    alpha=.8,
                    size=1)

g <- g + theme(legend.justification=c(1,1), legend.position=c(1,1))

g <- g + ggtitle("Standard Deviation and Standard Error of the distributions")

g <- g + scale_fill_discrete(
            breaks=c("exp", "avg"), 
            labels=c("of exponentials", "of averages")
        )

g
```

For the exponential distribution, it was plotted the first standard deviation
lines, which are $mean \pm \sqrt{var}$, and for the distribution of averages,
it was plotted the first standard errors, which are $mean \pm \sqrt{var}$.

The variance for the exponential distribution is 
`r s[s$type=='exp',]$var` and for the distribution of 
averages is `r s[s$type=='avg',]$var`. This difference is expected, as the
distribution of averages of exponentials are more concentrated to the mean of
the distribution.

When estimating the variance of the population, $\sigma^2$, the variance of 
the exponential distribution estimates it:

$S^2 = \sigma^2 = `r s[s$type=='exp',]$var`$.

Derivating the standard error formula:

$SE_{mean} = \frac{\sigma}{\sqrt{n}} \implies \sigma = SE_{mean} * \sqrt{n} = \sqrt{Var(\bar{X})*n} = \sqrt{`r s[s$type=='avg',]$var`*`r observations`} = `r s[s$type=='avg',]$var * observations`$

Each of the variances $\sigma^2$ is the square of the standard deviation, so,
respectively, `r sqrt(s[s$type=='exp',]$var)` and 
`r sqrt(s[s$type=='avg',]$var * observations)` are the
standard deviations from the exponential distribution and from the distribution
averages, whose are near the theoretical standard deviation of `r dist.sd`.


Distribution
------------
<!--
Via figures and text, explain how one can tell the distribution is approximately normal.
-->

```{r distribution, echo = FALSE, fig.height=3}
g <- ggplot(distributions[distributions$type == 'avg',], 
            aes(x=x, colour="averages"))

g <- g + geom_histogram( 
                        binwidth=.3, 
                        aes(y=..density..), 
                        colour="black",
                        fill="blue")

g <- g + stat_density(alpha=.8,
                      size=1,
                      geom="line",
                      position="identity")

g <- g + stat_density(data=std.normal, 
                      aes(x=x, colour="standard normal"), 
                      alpha = .8,
                      size=1, 
                      geom="line",
                      position="identity")

g <- g + scale_colour_manual(
            values=c("averages"="red", "standard normal"="green"),
            name="distribution")

g <- g + scale_linetype_manual(
            values=c("averages"=1, "standard normal"="dotted"),
            name="distribution")

g <- g + theme(legend.justification=c(1,1), legend.position=c(1,1))

g <- g + ggtitle("Distribution of Averages vs Standard Normal Distribution")

g
```

Plotting the densities for the distribution of averages of exponentials and for
a standard normal with the same median and standard deviation, we can see that
they are similar, so the Central Limit Theorem is verified, as expected.

# Appendix

Code for *Sample Mean versus Theoretical Mean*:

```{r ref.label='mean', results='hide', eval=FALSE}
```

Code for *Sample Variance versus Theoretical Variance*:

```{r ref.label='variance', results='hide', eval=FALSE}
```

Code for *Distribution*:

```{r ref.label='distribution', results='hide', eval=FALSE}
```
